---
title: "Methods overview"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

themis provides a large family of steps for dealing with imbalanced data. This article organizes them into a taxonomy, borrowed from [imbalanced-learn](https://imbalanced-learn.org/), so you can see how the methods relate to each other and choose between them. Every method is a recipe step, and most also have a direct-implementation function of the same name without the `step_` prefix.

```{r}
#| message: false
library(recipes)
library(themis)
```

## Over-sampling

Over-sampling grows the minority classes, either by replicating existing observations or by synthesizing new ones.

- **Random replication.** `step_upsample()` samples minority observations with replacement. It adds no new information but is fast and model-agnostic.
- **SMOTE family.** These synthesize new observations by interpolating between a minority observation and its nearest neighbors.
  - `step_smote()` is the original SMOTE for numeric predictors.
  - `step_bsmote()` (borderline-SMOTE) synthesizes only near the class boundary, where errors are most likely.
  - `step_svmsmote()` uses a support vector machine to locate the boundary before synthesizing.
  - `step_smotenc()` handles a mix of numeric and nominal predictors.
  - `step_smoten()` handles data sets where every predictor is nominal.
- **Adaptive.** `step_adasyn()` is like SMOTE but generates more synthetic points for the minority observations that are harder to learn (those surrounded by majority neighbors).
- **Smoothed bootstrap.** `step_rose()` generates a fully synthetic data set by drawing from smoothed kernels around the observations.

## Under-sampling

Under-sampling shrinks the majority classes. Most under-samplers in themis perform *prototype selection*, keeping a subset of the real rows. One, `step_cluster_centroids()`, performs *prototype generation*, creating new representative rows instead.

- **Random.** `step_downsample()` drops majority observations at random.
- **Distance / near-boundary.** These target observations by their position relative to the class boundary.
  - `step_nearmiss()` keeps the majority observations closest to the minority class.
  - `step_tomek()` removes the majority member of each Tomek link (a mutually-nearest pair from different classes).
  - `step_cnn()` (condensed nearest neighbors) keeps a subset that still classifies the data with a 1-nearest-neighbor rule.
  - `step_oss()` (one-sided selection) combines condensed nearest neighbors with Tomek links.
- **Neighborhood cleaning.** These remove observations that disagree with their neighborhood.
  - `step_enn()` (edited nearest neighbors) removes observations whose class differs from the majority of their neighbors. With `times` or `all_k` it becomes RENN or AllKNN.
  - `step_ncl()` (neighborhood cleaning rule) is a variant focused on cleaning the neighborhoods of minority observations.
- **Hardness.** `step_instance_hardness()` removes the majority observations that are hardest to classify, measured by how often their neighbors disagree.
- **Prototype generation.** `step_cluster_centroids()` summarizes each majority class with k-means cluster representatives: the centroids themselves with `voting = "soft"`, or the observation nearest each centroid with `voting = "hard"`.

## Combination

imbalanced-learn ships combination samplers (`SMOTEENN`, `SMOTETomek`) that over-sample and then clean the result. scikit-learn needs these as single objects because a pipeline cannot chain two samplers around one estimator. recipes has no such limitation, so the equivalents are ordinary step compositions.

```{r}
#| eval: false
# SMOTEENN: over-sample, then clean with edited nearest neighbors
recipe(class ~ ., data = train) |>
  step_smote(class) |>
  step_enn(class)

# SMOTETomek: over-sample, then remove Tomek links
recipe(class ~ ., data = train) |>
  step_smote(class) |>
  step_tomek(class)
```

Two things are worth knowing:

- **Order matters.** Over-sample first, then clean, so the cleaning step can remove the ambiguous synthetic points SMOTE tends to create near the boundary.
- **`step_enn()` cleans every class.** It flags any observation, including freshly synthesized minority points, whose class differs from its neighborhood. This matches imbalanced-learn's `SMOTEENN`, which configures its ENN stage with `sampling_strategy = "all"`. Standalone imbalanced-learn `EditedNearestNeighbours` instead defaults to cleaning "not minority", so themis's `step_enn()` is already oriented toward this combined use.

## Regression

Imbalance is not limited to classification. `step_smogn()` resamples an imbalanced *numeric* outcome, over-sampling rare regions of the response and under-sampling common ones, so themis applies to imbalanced regression as well.

## Choosing a method

- Start with `step_downsample()` or `step_upsample()` as fast baselines.
- Reach for the SMOTE family when synthesizing plausible minority observations helps, and pick the member that matches your predictor types (`step_smotenc()` / `step_smoten()` for nominal data).
- Use a cleaning under-sampler (`step_enn()`, `step_tomek()`, `step_ncl()`) when the majority class is noisy or overlaps the minority class, on its own or composed after SMOTE.
- Use `step_smogn()` for imbalanced regression.
