---
title: "How over_ratio and under_ratio work"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How over_ratio and under_ratio work}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)
```

```{r setup}
library(themis)
library(recipes)
library(dplyr)
```

## Introduction

Many of the sampling steps in themis accept an `over_ratio` or `under_ratio` argument that controls how much sampling is performed. This article explains what these arguments do and how to set them.

## Setup

To illustrate, we'll use a small data set with a known class imbalance. In this data set, class `"a"` has 100 observations, `"b"` has 65, and `"c"` has 20.

```{r}
set.seed(1)
imbalanced_data <- tibble(
  x = rnorm(185),
  class = factor(c(rep("a", 100), rep("b", 65), rep("c", 20)))
)

count(imbalanced_data, class)
```

## `over_ratio`

The `over_ratio` argument is used by [over-sampling steps](../reference/index.html#over-sampling). It controls the ratio of the minority-to-majority frequencies after sampling.

The target number of observations for each class is calculated as:

$$\text{target} = \lfloor \text{majority\_n} \times \text{over\_ratio} \rfloor$$

where $\text{majority\_n}$ is the number of observations in the most common class.

- **`over_ratio = 1` (default)**: All classes are upsampled to the frequency of the majority class, resulting in a perfectly balanced data set.

- **`over_ratio < 1`**: Minority classes are upsampled to a fraction of the majority class frequency, resulting in a partially balanced data set.

- **`over_ratio > 1`**: All classes, including the majority, are upsampled to the same target, resulting in a balanced data set that is larger than the original.

- If a class already has at least as many observations as the target, it is left unchanged.

### Examples

With the default `over_ratio = 1`, all classes are brought up to 100 observations (the size of the majority class):

```{r}
recipe(class ~ x, data = imbalanced_data) |>
  step_upsample(class, over_ratio = 1) |>
  prep() |>
  bake(new_data = NULL) |>
  count(class)
```

With `over_ratio = 0.5`, the target is `floor(100 * 0.5) = 50`. Only class `"c"` is upsampled to 50. Classes `"a"` and `"b"` already exceed the target and are left unchanged:

```{r}
recipe(class ~ x, data = imbalanced_data) |>
  step_upsample(class, over_ratio = 0.5) |>
  prep() |>
  bake(new_data = NULL) |>
  count(class)
```

With `over_ratio = 0.3`, the target is `floor(100 * 0.3) = 30`. Class `"c"` is upsampled from 20 to 30. Classes `"a"` and `"b"` both exceed the target and are left unchanged:

```{r}
recipe(class ~ x, data = imbalanced_data) |>
  step_upsample(class, over_ratio = 0.3) |>
  prep() |>
  bake(new_data = NULL) |>
  count(class)
```

## `under_ratio`

The `under_ratio` argument is used by [under-sampling steps](../reference/index.html#under-sampling). It controls the ratio of the majority-to-minority frequencies after sampling.

The target number of observations for each class is calculated as:

$$\text{target} = \lfloor \text{minority\_n} \times \text{under\_ratio} \rfloor$$

where $\text{minority\_n}$ is the number of observations in the least common class.

- **`under_ratio = 1` (default)**: All classes are downsampled to the frequency of the minority class, resulting in a perfectly balanced data set.

- **`under_ratio > 1`**: Majority classes are downsampled to a multiple of the minority class frequency, resulting in a partially balanced data set.

- **`under_ratio < 1`**: All classes, including the minority, are downsampled to the same target, resulting in a balanced data set that is smaller than the original.

- If a class already has at most as many observations as the target, it is left unchanged.

### Examples

With the default `under_ratio = 1`, all classes are brought down to 20 observations (the size of the minority class):

```{r}
recipe(class ~ x, data = imbalanced_data) |>
  step_downsample(class, under_ratio = 1) |>
  prep() |>
  bake(new_data = NULL) |>
  count(class)
```

With `under_ratio = 2`, the target is `floor(20 * 2) = 40`. Classes `"a"` and `"b"` are both downsampled to 40. Class `"c"` already has fewer than 40 observations and is left unchanged:

```{r}
recipe(class ~ x, data = imbalanced_data) |>
  step_downsample(class, under_ratio = 2) |>
  prep() |>
  bake(new_data = NULL) |>
  count(class)
```

With `under_ratio = 3`, the target is `floor(20 * 3) = 60`. Classes `"a"` and `"b"` are both downsampled to 60. Class `"c"` already has fewer than 60 observations and is left unchanged:

```{r}
recipe(class ~ x, data = imbalanced_data) |>
  step_downsample(class, under_ratio = 3) |>
  prep() |>
  bake(new_data = NULL) |>
  count(class)
```

## Choosing a ratio

Choosing the right ratio depends on your data and the model you are using. The default value of 1 gives a perfectly balanced data set, which is the most common choice. However, there are cases where a partial balance is preferable:

- **Preserving more majority class data**: A perfectly balanced data set from under-sampling discards most of the majority class data. Using `under_ratio > 1` retains more data at the cost of a less balanced class distribution.

- **Limiting over-sampling**: With a very large majority class, upsampling to perfect balance can generate a very large amount of synthetic data. Using `over_ratio < 1` limits the amount of synthetic data generated.

In practice, `over_ratio` and `under_ratio` are often treated as tunable hyperparameters and selected by cross-validation. See the [`dials`](https://dials.tidymodels.org/) package for the `over_ratio()` and `under_ratio()` parameter objects used in tuning.
